HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-08-03 scnn.tex
small modification

\citet{cohen2018spherical} uses the continuous spherical convolution.
The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is dominated by the two SHTs.
A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels.
Accelerated schemes exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
\citep{gorski2005healpix, reinecke2013libsharp}.
In comparison, filtering with our HealpixNet method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}).
% Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented by the highly optimized \pkg{libsharp} library used by HEALPix \citep{reinecke2013libsharp}.
% While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
% \TK{This I would leave out here, as this is a "spoiler" of the results section.}

Approaches by \citet{cohen2018spherical, boomsma2017spherical} can not be easily accelerated when the data lies on a part of the sphere only.
That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare.
One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals, on the latitude or longitude.
It is, however, not straightforward to identify those empty directions, and computations would still be wasted; consider, for example, a ring which mostly contains zeros but a few measurements.
With graphs however, the computation is done only for the used pixels.
This, however, results in some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}).
These can be mitigated by padding with zeros a small area around the measurements.

@@

The use of a graph to model the discretised sphere was also considered by~\citet{khasanova2017graphomni}.
The authors map omnidirectional images onto a sphere, with the goal of correcting for distortions induced by the lens, which was used for their acquisition.
Besides a different practical application, for which the method was designed, there are two important differences.
% First, a different sampling scheme is used: \cite{khasanova2017graphomni} uses a grid.
First, similarly to~\citet{boomsma2017spherical}, they use an equirectangular grid sampling.
% \TK{how is it different from us?}
Second, it uses a graph CNN based on a different parameterisation of the convolution kernel.
Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while the rotation invariance is desirable, the statistical layer used
% --- an operation which computes a set of spatially invariant statistics of the last feature map ---
% \{TK} this should be explained earlier
may not capture the information in an optimal way
\todo{and hinders performance (should be empirically verified)}.
\TK{This is a strong claim, can we support this?} \nati{It is not that strong as we use "may" right? We should probably soften it a bit. However the main idea is correct. If you compute some statistics on some features, you lose some information and may get worse results. }
Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest.
% While developing completely invariant transforms is a goal worth investigating, CNNs for image recognition, which are arguably the most researched and developed architectures, are not built to be fully invariant to translation.
% As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features.
% While that may not be ideal, it was shown to work well in practice.
% \TK{The two sentences above look like a tangent to me.. can we skip it an go to the main point, which is I think the fact that we use the stat layer?}

@@

>>>>

\citet{cohen2018spherical} uses the continuous spherical convolution.
The goal of this contribution was to define a rotation-equivariant convolution on the sphere, i.e., a rotation of the input implies the same rotation in the output.
The resulting convolution is performed by (i) a spherical harmonic transform (SHT), i.e., a projection on the spherical harmonics, (ii) a  multiplication in the spectral domain, and (iii) an inverse SHT. Hence the computational cost of a convolution is dominated by the two SHTs.
A naive implementation of the SHT costs $\bO(N_{pix}^2)$ operations, where $N_{pix}$ is the number of pixels.
Accelerated schemes exist for some sampling sets \citep[see][for examples]{mohlenkamp1999fast, rokhlin2006fast, reinecke2013libsharp}.
The main advantage of this approach is that it provides a mathematically well defined rotation-equivariant network. Nevertheless, even with accelerated SHTs, the convolutions remain expensive, limiting the practical use of this approach.
For example with the HEALPix sampling, which was designed to have a fast SHT by being iso-latitude, the computational cost of the SHT is $\bO(N_{pix}^{3/2}) = \bO(N_{side}^3) = \bO(\ell_{max}^3)$, where $\ell_{max}$ is the largest angular frequency\footnote{All pixels are placed on $N_{ring} = 4N_{side}-1 = \bO(\sqrt{N_{pix}})$ rings of constant latitude. Each ring has $\bO(\sqrt{N_{pix}})$ pixels. Thanks to this iso-latitude property, the SHT is computed using recurrence relations for Legendre polynomials on co-latitude and fast Fourier transforms (FFTs) on longitude. The computational cost is thus $\sqrt{N_{pix}}$ FFTs for a total cost of $\bO(N_{pix} \log(\sqrt{N_{pix}}))$, plus $\sqrt{N_{pix}}$ matrix multiplications of size $\sqrt{N_{pix}}$ for a total cost of $\bO(N_{pix}^{3/2})$.}
\citep{gorski2005healpix, reinecke2013libsharp}.
In comparison, filtering with our HealpixNet method scales as $\bO(N_{pix})$ (see \secref{efficient_convolution}).
% Moreover, \figref{filtering_speed} shows that a naive implementation of our method is more than ten times faster than the SHT implemented by the highly optimized \pkg{libsharp} library used by HEALPix \citep{reinecke2013libsharp}.
% While a comparison is theoretically possible, the experiments of~\cite{cohen2018spherical} are done using the geodesic grid sampling set and do not provide any code for the HEALPix one. Hence, a practical comparison is not possible.
% \TK{I would leave out here, as this is a "spoiler" of the results section.}

Approaches by \citet{cohen2018spherical, boomsma2017spherical} can not be easily accelerated when the data lies on a part of the sphere only.  \nati{Why is \citet{boomsma2017spherical} here? I think it can be trivally accelerated.}
That is an important use case in cosmology as measurements are often partial, i.e., full-sky maps are rare.
One could still fill the unseen part of the sphere with zeros, and avoid to compute empty integrals, on the latitude or longitude.
It is, however, not straightforward to identify those empty directions, and computations would still be wasted; consider, for example, a ring which mostly contains zeros but a few measurements.
With graphs however, the computation is done only for the used pixels.
This, however, results in some distortions due to border effects (see \figref{border_effects} and \ref{sec:border_effects}).
These can be mitigated by padding with zeros a small area around the measurements.

@@

The use of a graph to model the discretised sphere was also considered by~\citet{khasanova2017graphomni}.
The authors map omnidirectional images onto a sphere, with the goal of correcting for distortions induced by the lens, which was used for their acquisition.
Besides a different practical application, for which the method was designed, there are two important differences.
% First, a different sampling scheme is used: \cite{khasanova2017graphomni} uses a grid.
First, similarly to~\citet{boomsma2017spherical}, they use an equirectangular grid sampling.
% \TK{how is it different from us?}
Second, it uses a graph CNN based on a different parameterisation of the convolution kernel.
Moreover, they did not take advantage of a hierarchical pixelization of the sphere and resorted to dynamic pooling~\citep{kalchbrenner2014dcnn}.
Finally, while the rotation invariance is desirable, the statistical layer used
% --- an operation which computes a set of spatially invariant statistics of the last feature map ---
% \{TK} this should be explained earlier
may not capture the information in an optimal way
\todo{and hinders performance (should be empirically verified)}.
\TK{This is a strong claim, can we support this?} \nati{It is not that strong as we use "may" right? We should probably soften it a bit. However the main idea is correct. If you compute some statistics on some features, you lose some information and may get worse results. Nevertheless, since we are also using a statistical layer, I would juste remove this.}
Moreover, the chosen statistics have to be hand-crafted to capture statistics of interest.
% While developing completely invariant transforms is a goal worth investigating, CNNs for image recognition, which are arguably the most researched and developed architectures, are not built to be fully invariant to translation.
% As with the proposed spherical CNN, the fully connected layers at the end of the network depend on the spatial localization of features.
% While that may not be ideal, it was shown to work well in practice.
% \TK{The two sentences above look like a tangent to me.. can we skip it an go to the main point, which is I think the fact that we use the stat layer?}

@@

\subsection{Selected architecture}
\todo{Describe our fully convolutional architecture}
